MG7: Configurable and scalable 16S metagenomics data analysis -- new methods optimized for massive cloud computing

### 2015-09-22 docs/paper/out/paper.tex
Updated tex/pdf

The exponential growth of metagenomics is adding a significant plus of
complexity to the big data problem in genomics. In this new scenario
impacted by the wide scale and scope of the projects and by the
explosion of sequence data to be analyzed is especially opportune the
use of new possibilities that cloud computing approaches, new functional
and dependently typed programming languages and new database paradigms
as graph databases offer. To tackle the challenges of big data analysis
in this work we have used these new means to design and develop a new
open source methodology for analyzing metagenomics data, MG7. It
exploits the new possibilities that cloud computing offers to get a
system robust, programmatically configurable, modular, distributed,
flexible, scalable and traceable in which the biological databases of
reference sequences can be easily updated and/or frequently substituted
by new ones or by databases specifically designed for focused projects.
MG7 uses parallelization and distributed analysis based on AWS (Amazon
Web Services), with on-demand infrastructure as the basic paradigm and
allow the definition of complex workflows using a composable system for
scaling/parallelizing stateless computations designed for AWS that
counts with a static reproducible specification of dependencies and
behavior of the different components. The modelling of the taxonomy tree
is done using the new paradigm of graph databases of Bio4j that
facilitates the taxonomic assignment tasks and the calculation of the
taxa abundance values considering the hierarchic structure of taxonomy
tree. MG7 includes the new 16S-DB7 database built with a flexible and
sustainable system of updating and project-driven personalization.

@@

Genome researchers have raised the alarm over big data in the past
\citep{hayden2015genome}, but even a more serious challenge might be
faced with the metagenomics boom. If we compare metagenomics data with
other genomics data used in clinical genotyping we find a differential
feature: the key role of time. Thus, for example, in some longitudinal
studies, serial sampling from the same patient
\citep{faust2015metagenomics} along several weeks (or years) is being
used for the follow up of some intestinal pathologies, for studying the
evolution of the gut microbiome after antibiotic treatment, or for colon
cancer early detection
\citep[\citet{garrett2015cancer}]{zeller2014potential}. This need of
sampling across time adds more complexity to metagenomics data storage
and demands adapted algorithms to detect state variations across time as
well as idiosyncratic commonalities of the microbiome of each individual
\citep{franzosa2015identifying}. In addition to the intra-individual
sampling-time dependence, metagenomic clinical test results vary
depending on the specific region of extraction of the clinical specimen.
This local variability adds complexity to the analysis since different
localizations (different tissues, different anatomical regions, healthy
or tumour tissues) are required to have a sufficiently complete
landscape of the human microbiome. Moreover, re-analysis of old samples
using new tools and better reference databases might be also demanded
from time to time.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Sequence data:} the reads are larger, the sequencing depth and
  the number of samples of each project are considerably bigger. The
  first metagenomics studies were very local projects, while nowadays
  the most fruitful studies are done at a global level (international,
  continental, national). This kind of global studies has yielded the
  discovery of clinical biomarkers for diseases of the importance of
  cancer, obesity or inflammatory bowel diseases and has allowed
  exploring the biodiversity of varied earth environments.
\item
  \textbf{The genomics explosion:} its effect being felt in this case in
  the reference sequences. The immense amount of sequences available in
  public repositories demands new strategies for curation, update and
  storage of metagenomics reference databases: current models will
  (already) have problems to face the future avalanche of metagenomic
  sequence data.
\item
  \textbf{Cloud computing:} the appearance of new models for massive
  computation and storage such as the so-called cloud, or the widespread
  adoption of programming methodologies like functional programming, or,
  more speculatively, dependently typed programming. The new
  possibilities that these advances offer must have a direct impact in
  metagenomics data analysis.
\item
  \textbf{Open science:} the new social manner to do science,
  particularly so in genomics, brings its own set of requirements.
  Metagenomics evolves in a social and global scenario following a
  science democratization trend in which many small research groups from
  distant countries share a common big metagenomics project; this global
  cooperation demands systems allowing for reproducible data analysis,
  data interoperability, and tools and practices for asynchronous
  collaboration between different groups.
\end{enumerate}

>>>>

The exponential growth of metagenomics is adding a significant plus of
complexity to the big data problem in genomics. In this new scenario
impacted by the wide scale and scope of the projects and by the
explosion of sequence data to be analyzed is especially opportune the
use of new possibilities that cloud computing approaches, new functional
and dependently typed programming languages and new database paradigms
as graph databases offer. To tackle the challenges of big data analysis
in this work we have used these new means to design and develop a new
open source methodology for analyzing metagenomics data, MG7. It
exploits the new possibilities that cloud computing offers to get a
system robust, programmatically configurable, modular, distributed,
flexible, scalable and traceable in which the biological databases of
reference sequences can be easily updated and/or frequently substituted
by new ones or by databases specifically designed for focused projects.
MG7 uses parallelization and distributed analysis based on AWS (Amazon
Web Services), with on-demand infrastructure as the basic paradigm and
allow the definition of complex workflows using a composable system for
scaling/parallelizing stateless computations designed for AWS that
counts with a static reproducible specification of dependencies and
behavior of the different components. The modeling of the taxonomy tree
is done using the new paradigm of graph databases of Bio4j that
facilitates the taxonomic assignment tasks and the calculation of the
taxa abundance values considering the hierarchic structure of taxonomy
tree. MG7 includes the new 16S-DB7 database built with a flexible and
sustainable system of updating and project-driven personalization.

@@

Genome researchers have raised the alarm over big data in the past
\citep{hayden2015genome}, but even a more serious challenge might be
faced with the metagenomics boom. If we compare metagenomics data with
other genomics data used in clinical genotyping we find a differential
feature: the key role of time. Thus, for example, in some longitudinal
studies, serial sampling from the same patient
\citep{faust2015metagenomics} along several weeks (or years) is being
used for the follow up of some intestinal pathologies, for studying the
evolution of the gut microbiome after antibiotic treatment, or for colon
cancer early detection
\citep[\citet{garrett2015cancer}]{zeller2014potential}. This need of
sampling across time adds more complexity to metagenomics data storage
and demands adapted algorithms to detect state variations across time as
well as idiosyncratic commonalities of the microbiome of each individual
\citep{franzosa2015identifying}. In addition to the intra-individual
sampling-time dependence, metagenomic clinical test results vary
depending on the specific region of extraction of the clinical specimen.
This local variability adds complexity to the analysis since different
localizations (different tissues, different anatomical regions, healthy
or tumor tissues) are required to have a sufficiently complete landscape
of the human microbiome. Moreover, re-analysis of old samples using new
tools and better reference databases might be also demanded from time to
time.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Sequence data:} the reads are larger, the sequencing depth and
  the number of samples of each project are considerably bigger. The
  first metagenomics studies were very local projects, while nowadays
  the most fruitful studies are done at a global level (international,
  continental, national). This kind of global studies has yielded the
  discovery of clinical biomarkers for diseases of the importance of
  cancer, obesity or inflammatory bowel diseases and has allowed
  exploring the biodiversity of varied earth environments.
\item
  \textbf{The genomics explosion:} its effect being felt in this case in
  the reference sequences. The immense amount of sequences available in
  public repositories demands new strategies for curation, update and
  storage of metagenomics reference databases: current models will
  (already) have problems to face the future avalanche of metagenomic
  sequence data.
\item
  \textbf{Cloud computing:} the appearance of new models for massive
  computation and storage such as the cloud-based platforms, or the
  widespread adoption of programming methodologies like functional
  programming, or, more speculatively, dependently typed programming.
  The new possibilities that these advances offer must have a direct
  impact in metagenomics data analysis.
\item
  \textbf{Open science:} the new social manner to do science,
  particularly so in genomics, brings its own set of requirements.
  Metagenomics evolves in a social and global scenario following a
  science democratization trend in which many small research groups from
  distant countries share a common big metagenomics project; this global
  cooperation demands systems allowing for reproducible data analysis,
  data interoperability, and tools and practices for asynchronous
  collaboration between different groups.
\end{enumerate}
